---
title: "Getting started with llegir"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with llegir}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA
)

# data/ and output/ paths in this vignette are relative to the package root,
# so evaluate chunks there rather than in vignettes/
knitr::opts_knit$set(root.dir = normalizePath('..'))

# data/CSF_Myeloid_hdWGCNA.rds is a large, gitignored/buildignored dev-only
# fixture (see .Rbuildignore) -- it exists on a dev machine but never inside
# the tarball R CMD check builds and runs against, so every chunk below is
# skipped there instead of erroring on a missing file
data_available <- file.exists('data/CSF_Myeloid_hdWGCNA.rds')
knitr::opts_chunk$set(eval = data_available)
```

```{r, echo = FALSE, eval = TRUE, results = 'asis'}
if (!data_available) {
    cat(
        "*This vignette's code chunks require a local ",
        "`data/CSF_Myeloid_hdWGCNA.rds` dev fixture (gitignored and build-ignored, ",
        "not shipped with the package) and are skipped here. ",
        "The prose below still walks through the full workflow.*\n",
        sep = ''
    )
}
```

**llegir** is an R package to automate the interpretation of gene modules in a specific context. From an input gene expression dataset (single-cell, spatial transcriptomics, or bulk RNA-seq) and a set of gene modules, **llegir** gathers a standardized bundle of evidence per module using a pre-built toolbox of analysis functions, and drafts an evidence-backed interpretation via a LLM-enabled synthesis layer.

*NOTE* this package is at an experimental development stage, and much of the interface is subject to change.

## Setup data and define the `ModuleSet`

In this vignette, we demonstrate **llegir** using a scRNA-seq dataset of myeloid cells from cerebrospinal fluid (CSF) cells profiled across several CNS conditions, with co-expression modules computed by hdWGCNA. 

To start the **llegir** pipeline, we first define the `ModuleSet`, which contains the following attributes: `modules()`, `gene_membership()`, `module_scores()`, `expression()`, `metadata()`, `capabilities()`, `pkg_versions()`. The function `hdWGCNA_ModuleSet()` automatically formats the `ModuleSet` object for hdWGCNA results, but this framework is flexible to modules from other packages.

```{r, message = FALSE, warning = FALSE, purl = FALSE}
library(Seurat)
library(hdWGCNA)
library(tidyverse)
library(llegir)

# load the CSF dataset, which has already been processed with hdWGCNA
seurat_obj <- readRDS('data/CSF_Myeloid_hdWGCNA.rds')

# initialize the ModuleSet 
ms <- hdWGCNA_ModuleSet(seurat_obj)
modules(ms)

```

A quick look at the two attributes most tools lean on: the ranked
gene membership of a module and the cell-level metadata.

```{r, purl = FALSE}
head(gene_membership(ms, 'MM1'))
```

```{r, purl = FALSE}
head(metadata(ms))
```

## Global experiment grounding

Before diving into individual modules, **llegir** first gathers global context about the whole experiment, once, via a small toolbox of **dataset tools** -- the same kind of standardized, evidence-emitting functions as the per-module tools below, just scoped to the whole `ModuleSet` instead of one module. `dataset_composition_tool()` summarizes cell-state census and covariate balance (e.g. is one cell state skewed across diagnosis groups?), and `dataset_variance_structure_tool()` regresses each principal component of the pseudo-bulk expression matrix against declared metadata covariates, so synthesis can tell whether a module's cross-condition signal reflects real biology or a technical axis (batch, sequencing depth) riding the same PC.

`run_dataset_context()` mirrors `run_orchestrator()` below, but runs its tool list once against the whole `ModuleSet` rather than once per module:

```{r, purl = FALSE}
dataset_tool_config <- list(
    list(fn = dataset_composition_tool, params = list(group_col = 'lv2_annot', condition_col = 'diagnosis')),
    list(fn = dataset_variance_structure_tool, params = list(
        covariates = c('diagnosis', 'nUMI'), condition_col = 'diagnosis'
    ))
)

dataset_ctx <- run_dataset_context(ms, dataset_tool_config)
dataset_ctx$dataset_fragments[[1]]$compact_summary
```

`dataset_variance_structure_tool()` needs a pseudo-bulk view of the data (see `pseudobulk_ModuleSet()` / `with_pseudobulk()`) to regress against, so it skips gracefully -- rather than erroring -- when `ms` doesn't carry one, as here. The resulting `dataset_ctx` is passed into `run_synthesis_orchestrator()` (or `synthesize_module()`) alongside `desc`, via `dataset_context = dataset_ctx`, injecting a DATASET CONTEXT block into every module's synthesis prompt (see the `## Interpreting evidence` section below).

## Orchestrating a chain of analysis tools

Each module is characterized by running a set of analysis tools. Every tool takes the same inputs (the `ModuleSet`, a module, and some parameters, bundled together as `ctx`) and returns one `evidence_fragment`: a standardized summary of what that tool found. To demonstrate this, we run a single tool on its own to see what it produces:

```{r, purl = FALSE}

# set up the context (ctx) for the tool
ctx <- list(ms = ms, module_id = 'MM1', params = list(n_hubs = 10))

# run the top genes tool to get the top 10 genes by membership
top_frag <- top_genes_tool(ctx)

# sanity check:
validate_evidence_fragment(top_frag)

# show the result
top_frag$compact_summary
```

**llegir** ships several pre-built analysis tools:

- `top_genes_tool()`: top genes by kME
- `cluster_dme_tool()`: differential module expression across clusters
- `geneset_enrichment_tool()`: GO/pathway enrichment
- `signature_correlation_tool()`: co-variation with a signature library

Additionally, external results you have already computed elsewhere can be leveraged the same way via `import_fragment()`.

We run a chain of tools over the module set using `run_orchestrator()`, which then collects the results into one **evidence packet** per module, saved as a JSON file. The tool list pairs each tool (`fn`) with its parameters, and `modules_use` limits the run to three modules for this example:

```{r, purl = FALSE}
modules_use <- c('MM1', 'MM2', 'MM3')

tool_config <- list(
    list(fn = top_genes_tool, params = list(n_hubs = 25)),
    list(fn = cluster_dme_tool, params = list(group_by = 'lv2_annot')),
    list(fn = geneset_enrichment_tool, params = list(
        n_hubs = 25,
        db_files = c(
            GO_BP = 'data/GO_Biological_Process_2026.txt',
            Hallmark = 'data/h.all.v2026.1.Hs.symbols.gmt'
        )
    ))
)

packets <- run_orchestrator(ms, tool_config, output_dir = 'output/evidence_packets', modules_use = modules_use)
```

Each packet holds its evidence fragments plus a `packet_hash`, a fingerprint of
the evidence it contains. The same evidence always produces the same hash, which
makes runs easy to track and reproduce.

```{r, purl = FALSE}
packet <- packets[['MM1']]
packet$packet_hash
length(packet$fragments)
```

## Provide dataset context

Before interpreting the modules, we provide a brief description of the dataset `dataset_description()`. This biological context is given to the model so it interprets each module in an appropriate setting (the same gene program can mean something different in CSF myeloid cells than in a solid tumor). It is required: **llegir** will not run synthesis without it.

```{r, purl = FALSE}
desc <- dataset_description(
    species = 'human',
    tissue = 'cerebrospinal fluid (CSF)',
    cell_compartment = 'myeloid cells',
    assay = 'single-cell RNA-seq (10x)',
    conditions = c(
        'Glioblastoma', 'Brain Metastasis', 'Primary CNS lymphoma',
        'Secondary CNS lymphoma', 'Inflammatory / other neuroinflammatory'
    ),
    notes = 'Modules are CSF-myeloid co-expression programs; interpret in a CNS-myeloid, neuro-oncology / neuroinflammation context.'
)
cat(render_dataset_description(desc))
```

## Interpreting evidence

`run_synthesis_orchestrator()` turns the evidence packets into written interpretations. Each module's evidence packet is sent to a user-specified LLM, checks that every claim is actually backed by the evidence (`enforce_faithfulness()`), and combines the model's confidence with the evidence-based confidence (`fuse_confidence()`). It saves an interpretation and a readable Markdown paragraph per module, plus a `review_queue.tsv` and `manifest.json` for the whole run.

The language model's role is deliberately limited: it only ever sees the summarized evidence, never the raw data tables, and it never runs any analysis itself.

You can use different model providers through `resolve_backend()` (here
`'github'`; `'gemini'` also works), which reads your API key from the
environment (e.g. `GITHUB_PAT` or `GEMINI_API_KEY`). Wrapping it in
`cached_backend()` saves each result to disk, so re-running synthesis on
unchanged packets doesn't spend another API call.

```{r, purl = FALSE}
backend <- cached_backend(
    resolve_backend('github'),
    provider = 'github',
    model = 'gpt-4o-mini',
    prompt_template_version = PROMPT_TEMPLATE_VERSION
)

interps <- run_synthesis_orchestrator(
    packets, desc, backend = backend, output_dir = 'output/interpretations', dataset_context = dataset_ctx
)
```

Inspect one interpretation: its proposed label, fused confidence, and the
rendered paragraph.

```{r, purl = FALSE}
interp <- interps[['MM1']]
interp$proposed_label
interp$confidence$score
cat(render_paragraph(interp))
```

The review queue (`build_review_queue()`) collects only the interpretations
flagged for a closer look (`needs_review()`), sorted lowest-confidence first,
so you know which modules to check by hand:

```{r, purl = FALSE}
build_review_queue(interps)
```
