---
title: "Getting started with sentit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with sentit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)
```

**sentit** ("to sense", Catalan) is a Module Interpretation Engine: it loads a
gene co-expression module object (starting with hdWGCNA), gathers a
standardized bundle of evidence per module through a toolbox of deterministic
R functions, and drafts an evidence-backed interpretation paragraph via a
model-agnostic synthesis layer with faithfulness checking and confidence
fusion.

This vignette walks the whole pipeline end to end on a small, self-contained
synthetic dataset -- no Seurat/hdWGCNA object, no external data file, and no
network access or API key required. It builds entirely offline via
[`mock_backend()`], so it runs the same way in `R CMD check` as it does
interactively.

*sentit is experimental: the public API, especially the `ModuleSet` adapter
contract and the custom-tool interface, is expected to evolve before a 1.0
release.*

## 1. A `ModuleSet`

Every core tool depends only on the `ModuleSet` adapter contract --
[`modules()`], [`gene_membership()`], [`module_scores()`], [`expression()`],
[`metadata()`], [`pkg_versions()`] -- never on a specific backend package.
[`hdWGCNA_ModuleSet()`] is the real-world adapter, built from a Seurat object
that has been run through hdWGCNA. For this vignette we instead use
[`sentit_example_moduleset()`], a small simulated dataset (two co-expressed
gene modules plus background noise genes, 200 cells) that satisfies the same
contract, so every tool below runs unmodified against it.

```{r}
library(sentit)

ms <- sentit_example_moduleset()
modules(ms)
```

```{r}
head(gene_membership(ms, 'module_a'))
```

```{r}
head(metadata(ms))
```

`module_a` is simulated to associate with the `diagnosis` column, and
`module_b` with `cell_type` -- used below to show a real (not null) evidence
signal.

## 2. Evidence tools

Each tool takes a `ctx <- list(ms, module_id, params)` and returns a single,
validated `evidence_fragment` -- see [`evidence_fragment()`] and the shipped
`evidence_fragment.schema.json` for the full contract.

```{r}
ctx <- list(ms = ms, module_id = 'module_a', params = list(n_hubs = 10))
hub_frag <- hub_genes_tool(ctx)
validate_evidence_fragment(hub_frag)
hub_frag$compact_summary
```

```{r}
meta_frag <- module_by_metadata_tool(list(
    ms = ms, module_id = 'module_a',
    params = list(column = 'diagnosis', column_type = 'categorical')
))
meta_frag$compact_summary
```

```{r}
cluster_frag <- cluster_dme_tool(list(
    ms = ms, module_id = 'module_b',
    params = list(group_by = 'cell_type')
))
cluster_frag$compact_summary
```

A fourth core tool, [`geneset_enrichment_tool()`], runs offline GO/pathway
enrichment against a local GMT file (`ctx$params$db_files`); it's omitted
here since no gene-set library ships with the package, but it follows the
same `ctx -> evidence_fragment` contract as the tools above.

User-supplied result tables (e.g. a DME table computed outside sentit) can be
folded into the same pipeline via [`import_fragment()`] /
[`import_fragment_tool()`], tagged `provenance$source = 'user_supplied'`
rather than `'computed'`.

## 3. Evidence packets

[`run_module()`] runs a configured list of tools against one module and
bundles the fragments into a validated, hashed evidence packet --
[`build_evidence_packet()`] under the hood. The `packet_hash` is a content
hash of the fragments (timestamps excluded), so it's a reproducibility
fingerprint: the same evidence always hashes the same way.

```{r}
tool_config <- list(
    list(fn = hub_genes_tool, params = list(n_hubs = 10)),
    list(fn = module_by_metadata_tool, params = list(column = 'diagnosis', column_type = 'categorical'))
)
packet <- run_module(ms, 'module_a', tool_config, input_hash = 'vignette-example')
packet$packet_hash
length(packet$fragments)
```

[`run_orchestrator()`] does this for every module in a `ModuleSet` at once,
writing one packet JSON per module (plus, optionally, every fragment's full
result table via `tables_dir`):

```{r}
out_dir <- tempfile('packets_')
packets <- run_orchestrator(ms, tool_config, output_dir = out_dir)
list.files(out_dir)
```

A single module failing (e.g. a degenerate module tripping up one tool)
doesn't take down the whole batch -- it's warned and recorded as `NULL`,
leaving every other module's packet intact.

## 4. Dataset context

Synthesis requires a [`dataset_description()`] -- the biological context
prepended to every prompt, so the model interprets a module in the right
frame (e.g. CSF myeloid vs. tumor tissue changes what a given program means).
It's a hard requirement, not a default: [`validate_dataset_description()`]
errors on a missing/empty required field.

```{r}
desc <- dataset_description(
    species = 'human', tissue = 'simulated tissue', cell_compartment = 'simulated cells',
    assay = 'simulated scRNA-seq', conditions = c('case', 'control')
)
cat(render_dataset_description(desc))
```

## 5. Synthesis

[`synthesize_module()`] runs the full packet -> interpretation pipeline:
calls a backend via [`synthesize_interpretation()`], enforces citation
faithfulness ([`enforce_faithfulness()`] -- every claim's `fragment_ids` must
actually exist in the packet, and its `direction` must match), and fuses
model and deterministic confidence ([`fuse_confidence()`]).

The model's role is bounded by design: it only ever sees the compact prompt
built by [`build_system_prompt()`] / [`build_user_prompt()`] (fragment
summaries and top findings, never raw result tables), and it never runs
analysis code.

Here, [`mock_backend()`] stands in for a live model -- a canned, fixed
response that never touches the network, used by the whole test suite and by
this vignette:

```{r}
interp <- synthesize_module(packet, desc, backend = mock_backend())
interp$proposed_label
interp$confidence$score
```

```{r}
cat(render_paragraph(interp))
```

### Live backends

sentit's synthesis backends are pluggable via [`resolve_backend()`]
(`provider = 'github'` or `'gemini'`, both built on [`ellmer_backend()`]) or
[`ellmer_backend()`] directly with any `ellmer::chat_*()` constructor. These
need network access and a provider API key (picked up by `ellmer` itself from
the provider's environment variable, e.g. `GITHUB_PAT` or `GEMINI_API_KEY`),
so they aren't exercised in this vignette or in `R CMD check` -- only shown
here for reference:

```{r, eval = FALSE}
backend <- resolve_backend(provider = 'gemini')
interp <- synthesize_module(packet, desc, backend = backend)
```

[`cached_backend()`] wraps any backend with an on-disk cache keyed on the
packet hash + provider + model + prompt template version, so re-running
synthesis over an unchanged packet never re-spends an API call.

## 6. Batch synthesis: review queue and manifest

[`run_synthesis_orchestrator()`] runs [`synthesize_module()`] over a batch of
packets (e.g. the output of [`run_orchestrator()`]), writing per-module
interpretation JSON and rendered Markdown paragraphs, plus a run-level
`review_queue.tsv` and `manifest.json`:

```{r}
synth_dir <- tempfile('synthesis_')
interps <- run_synthesis_orchestrator(packets, desc, backend = mock_backend(), output_dir = synth_dir)
list.files(synth_dir)
```

The review queue ([`build_review_queue()`]) lists only flagged
interpretations ([`needs_review()`]) -- a triage queue for a human reviewer,
not a summary of every module -- sorted lowest-confidence first:

```{r}
build_review_queue(interps)
```

The mock backend always returns the same fixed, generic content regardless
of the actual evidence packet -- so its citations and confidence score are
never truly grounded in this vignette's per-module evidence, and
`needs_human_review` is expected here. A live model, synthesizing from the
real packet contents, will typically agree with the deterministic evidence
more often and flag fewer modules.
