---
title: "Using a local open-source LLM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using a local open-source LLM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This vignette assumes you have already read *Getting started with llegir* and run its core workflow. The only thing that changes here is the backend. Everything from `run_orchestrator()` onward is identical.

## Why run locally?

**llegir** synthesis calls `chat$chat_structured()` and expects the model to return JSON that validates against the `interpretation` schema. That means the serving engine must support **grammar-constrained ("guided") decoding** — a model that can chat freely but cannot reliably emit schema-constrained JSON will fail `validate_interpretation()` every time. vLLM provides mature, well-tested guided decoding via its OpenAI-compatible API, which is exactly what `ellmer::chat_vllm()` targets. Everything downstream — faithfulness checking, confidence fusion, caching, rendering — is unchanged.

## Choosing a model

Pick an **instruct-tuned** model from a family with strong structured-output behavior. The **Qwen instruct** series (e.g. `Qwen/Qwen2.5-7B-Instruct`) is a reliable starting point: it fits comfortably on most modern data-center GPUs, serves fast, and handles schema-constrained output well. Start at 7B regardless of available VRAM; once the full pipeline works, swapping the model id for a 14B or 32B checkpoint is a one-line change. Avoid base (non-instruct) checkpoints and "reasoning" models — the latter emit chain-of-thought that conflicts with strict JSON decoding.

## Serving the model with vLLM

The commands below run on a **GPU node**. On an HPC cluster they belong inside a GPU job allocation; scheduler headers and connection recipes (SSH tunnel, cross-node URL) vary by site — see `docs/local_llm_backend.md` for the full setup guide and a validated CNAG example.

**One-time setup (run on a login node that has internet access):**

```{bash}
micromamba create -n vllm python=3.11 -y
micromamba activate vllm
pip install vllm "huggingface_hub[cli]"

export HF_HOME=/path/to/scratch/hf_cache   # large, shared, GPU-node-visible
hf download Qwen/Qwen2.5-7B-Instruct
# gated models (some Llama/Gemma repos) also need: hf auth login
```

**Each run (inside the GPU allocation):**

```{bash}
export HF_HOME=/path/to/scratch/hf_cache
export HF_HUB_OFFLINE=1          # if compute nodes have no outbound internet
export VLLM_API_KEY=EMPTY

vllm serve Qwen/Qwen2.5-7B-Instruct \
    --served-model-name qwen \
    --host 127.0.0.1 \
    --port 8000 \
    --max-model-len 8192 \
    --gpu-memory-utilization 0.90
```

`--served-model-name qwen` is the short alias clients use — this is the `model` value you pass to **llegir**. Wait for `Application startup complete` before running R.

## Connecting llegir

Set the server URL, then call `local_backend()`. That is the only change from the getting-started workflow.

```{r}
# optional: point llegir at a non-default URL (defaults to http://127.0.0.1:8000/v1)
Sys.setenv(LLEGIR_LLM_URL = 'http://127.0.0.1:8000/v1')
```

`llm_server_status()` confirms connectivity and returns the list of served models:

```{r}
library(llegir)

status <- llm_server_status()
# $base_url: "http://127.0.0.1:8000/v1"
# $models:   "qwen"
```

`local_backend()` combines the connectivity check, model auto-discovery, and `cached_backend()` into a single call:

```{r}
backend <- local_backend()
```

If more than one model is served, pass the alias explicitly:

```{r}
backend <- local_backend(model = 'qwen')
```

From here the synthesis call is **identical to getting-started** — the `backend` line above is the only difference:

```{r}
interps <- run_synthesis_orchestrator(
    packets, desc, backend = backend,
    output_dir = 'output/interpretations',
    dataset_context = dataset_ctx
)

interp <- interps[['MM1']]
interp$proposed_label
interp$confidence$score
cat(render_paragraph(interp))
```

## What's next

Everything downstream — `build_review_queue()`, `write_synthesis_manifest()`, `export_agent_workspace()` — is identical regardless of which backend produced the interpretations. For HPC scheduler setup, cross-node URL configuration, SSH tunnels, quantized checkpoints, and troubleshooting, see `docs/local_llm_backend.md`.
