---
title: 'llegir interactive workflow test'
output: html_document
---

Source/run this line-by-line in an interactive R session to walk the core
package workflow end-to-end on three modules (MM1-MM3). `conda activate
hdWGCNA` first, then step through each chunk and inspect the intermediate
objects in the console as prompted.

```{r setup}
#---------------------------------------------------------
# setup environment and load the dataset
#---------------------------------------------------------

devtools::load_all()

# load the seurat object
data_path <- 'data/CSF_Myeloid_hdWGCNA.rds'
seurat_obj <- readRDS(data_path)

# define the set of modules to use for this test
modules_use <- c('MM1', 'MM2', 'MM3')
```

```{r custom-tool}
#---------------------------------------------------------
# define a custom, script-local evidence tool (Milestone 2 registry demo)
#---------------------------------------------------------

# this is a fully user-defined tool that lives outside the package. it obeys the
# same contract as the package-native tools: it takes a single tool ctx
# (list(ms, module_id, params), built by run_orchestrator) and returns a
# validated evidence_fragment. `ms` and `n_hubs` arrive via ctx$ms /
# ctx$params$n_hubs -- surfaced exactly like top_genes_tool does it -- so this
# slots straight into tool_config with no orchestrator changes.

# small helper: hit the NCBI E-utils gene endpoints for one symbol and return a
# one-line human summary. esearch resolves the human gene symbol to a UID, then
# esummary pulls the curated description. wrapped by the caller in tryCatch so a
# network failure degrades to the offline mock rather than aborting the run.
.ncbi_summary_one <- function(symbol){

    Sys.sleep(0.4)

    esearch_url <- paste0(
        'https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi',
        '?db=gene&retmode=json&term=', utils::URLencode(symbol),
        '%5Bsym%5D+AND+human%5Borgn%5D'
    )
    uid <- jsonlite::fromJSON(esearch_url)$esearchresult$idlist[1]
    if (is.na(uid) || length(uid) == 0) return(NA_character_)

    esummary_url <- paste0(
        'https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esummary.fcgi',
        '?db=gene&retmode=json&id=', uid
    )
    rec <- jsonlite::fromJSON(esummary_url)$result[[uid]]

    # NCBI's short blurb ('summary') is richer but often empty; fall back to the
    # one-line 'description' so every resolved gene still gets a non-empty string
    summary_text <- rec$summary
    if (is.null(summary_text) || !nzchar(summary_text)) summary_text <- rec$description
    summary_text %||% NA_character_
}

# vectorised lookup with an offline-safe fallback: try the live E-utils call for
# the whole gene vector, and if anything fails (no network, throttling, budget
# lockdown) synthesise a deterministic placeholder so the tool still emits a
# well-formed fragment offline
ncbi_gene_summaries <- function(genes){
    live <- tryCatch(
        vapply(genes, .ncbi_summary_one, character(1)),
        error = function(e) NULL
    )
    if (is.null(live)) {
        return(paste0('[mock summary] ', genes, ': human protein-coding gene (offline placeholder).'))
    }

    # patch only the genes E-utils could not resolve, keep real hits intact
    live[is.na(live)] <- paste0(
        '[mock summary] ', genes[is.na(live)], ': no NCBI record resolved (offline placeholder).'
    )
    unname(live)
}

# the tool proper -- same shape as top_genes_tool: rank the module's genes by
# kME, keep the top n_hubs, and attach an NCBI one-liner to each
ncbi_gene_summary_tool <- function(ctx){
    n_hubs <- ctx$params$n_hubs %||% 5

    gm <- gene_membership(ctx$ms, ctx$module_id)
    top <- utils::head(gm, n_hubs)
    top$summary <- ncbi_gene_summaries(top$gene_name)

    top_findings <- lapply(seq_len(nrow(top)), function(i){
        list(gene = top$gene_name[i], kme = top$kme[i], summary = top$summary[i])
    })

    compact_summary <- paste0(
        'NCBI summaries for top ', nrow(top), ' hub genes: ',
        paste(top$gene_name, collapse = ', ')
    )

    evidence_fragment(
        fragment_id = 'ncbi_gene_summary',
        tool_id = 'ncbi_gene_summary',
        module_id = ctx$module_id,
        type = 'ranked_genes',
        result = top,
        compact_summary = compact_summary,
        top_findings = top_findings,
        effect_strength = if (nrow(top) > 0) max(top$kme) else 0,
        direction = 'na',
        provenance = make_provenance(
            tool_version = '0.1',
            params = list(n_hubs = n_hubs),
            pkg_versions = pkg_versions(ctx$ms)
        )
    )
}

```

```{r moduleset}

# initialize the ModuleSet adapter
ms <- hdWGCNA_ModuleSet(seurat_obj)
modules(ms)

# run the capabilities check
capabilities(ms)

# smoke-test the script-local custom tool standalone on MM1 before wiring it into
# the orchestrator -- inspect result / top_findings and confirm each gene carries
# a summary string (live NCBI text, or a '[mock summary]' placeholder if offline)
ncbi_demo_frag <- ncbi_gene_summary_tool(list(ms = ms, module_id = 'MM1', params = list(n_hubs = 5)))
ncbi_demo_frag$result
ncbi_demo_frag$top_findings
validate_evidence_fragment(ncbi_demo_frag)
```

## Global Dataset-Level Grounding

Dataset tools mirror the per-module tools above but run once against the
whole `ModuleSet`, via `run_dataset_context()`, before any per-module tool
call -- the global framing every module's synthesis prompt should see
before it interprets a module as biology rather than, say, cell-type
imbalance or a batch effect.

```{r dataset-context}
#---------------------------------------------------------
# global dataset-level grounding (runs once, before the per-module loop)
#---------------------------------------------------------

# dataset_composition_tool checks cell-state balance against a condition
# covariate; dataset_variance_structure_tool regresses each pseudo-bulk PC
# against declared covariates so synthesis can tell real biology from a
# technical (batch/depth) axis riding the same PC. variance_structure needs
# a pseudo-bulk view (see pseudobulk_view()), which ms doesn't carry yet at
# this point in the script, so it skips gracefully here -- it is re-run
# below once with_pseudobulk() attaches one
dataset_tool_config <- list(
    list(fn = dataset_composition_tool, params = list(group_col = 'lv2_annot', condition_col = 'diagnosis')),
    list(fn = dataset_variance_structure_tool, params = list(
        covariates = c('diagnosis', 'nUMI'), condition_col = 'diagnosis'
    ))
)

dataset_ctx <- run_dataset_context(ms, dataset_tool_config)

# inspect which fragments actually landed -- only composition_summary at this
# point, since variance_structure had nothing to compute against yet
vapply(dataset_ctx$dataset_fragments, function(f) f$tool_id, character(1))
dataset_ctx$dataset_fragments[[1]]$compact_summary
```

```{r core-tools}

#---------------------------------------------------------
# run core evidence tools over MM1-MM3 only
#---------------------------------------------------------

hallmark_gmt <- c(Hallmark = 'data/h.all.v2026.1.Hs.symbols.gmt')

# define the list of tools to run -- the last entry is the script-local
# ncbi_gene_summary_tool injected right alongside the package-native tools, which
# is all it takes for the registry/orchestrator to treat it as a first-class tool
tool_config <- list(
    list(fn = top_genes_tool, params = list(n_hubs = 30)),
    list(fn = cluster_dme_tool, params = list(group_by = 'lv2_annot')),
    list(fn = geneset_enrichment_tool, params = list(
        n_hubs = 25,
        db_files = c(GO_BP = 'data/GO_Biological_Process_2026.txt', hallmark_gmt)
    )),
   # list(fn = signature_correlation_tool, params = list(library_files = hallmark_gmt)), # commented this out for now because it takes a long time to run
    list(fn = ncbi_gene_summary_tool, params = list(n_hubs = 5))
)

# run the orchestrator over the list of tools to construct evidence packets per module
packets <- run_orchestrator(ms, tool_config, output_dir = 'output/evidence_packets', modules_use = modules_use)

# inspect one packet's fragments before moving on
packets[['MM1']]
packets[['MM1']]$fragments[[1]]

#---------------------------------------------------------
# run synthesis on the same three modules via GitHub Models
#---------------------------------------------------------

desc <- dataset_description(
    species = 'human',
    tissue = 'cerebrospinal fluid (CSF)',
    cell_compartment = 'myeloid cells',
    assay = 'single-cell RNA-seq (10x)',
    conditions = c(
        'Glioblastoma', 'Brain Metastasis', 'Primary CNS lymphoma',
        'Secondary CNS lymphoma', 'Inflammatory / other neuroinflammatory'
    ),
    notes = 'Modules are CSF-myeloid co-expression programs; interpret in a CNS-myeloid, neuro-oncology / neuroinflammation context.'
)

backend <- cached_backend(
    resolve_backend('github'), provider = 'github', model = .default_models[['github']],
    prompt_template_version = PROMPT_TEMPLATE_VERSION
)

interps <- run_synthesis_orchestrator(
    packets, desc, backend, output_dir = 'output/interpretations', dataset_context = dataset_ctx
)

# inspect one interpretation
interps[['MM1']]
needs_review(interps[['MM1']])
```

```{r user-weights-demo}
#---------------------------------------------------------
# demonstrate tool importance overrides via user_weights (docs/milestones/
# milestone_fused_confidence.md S4) -- budget: stays to the one already-run
# MM1 packet/module, no extra synthesis calls beyond this single re-run
#---------------------------------------------------------

# MM1's packet carries the three core fragments (top_genes, cluster_dme,
# geneset_enrichment) plus the script-local ncbi_gene_summary_tool fragment;
# this is calculate_fusion_score() with the default structural tiers -- the
# same E_pool/E_evidence already baked into interps[['MM1']]$confidence
mm1_packet <- packets[['MM1']]
default_fusion <- calculate_fusion_score(mm1_packet$fragments)
default_fusion$matrix

# mute the script-local ncbi_gene_summary tool entirely (u = 0 -- it isn't
# registered via register_tool(), so it would otherwise fall back to the
# medium tier by default) and boost geneset_enrichment, structurally the
# low tier, up toward high -- e.g. a user who trusts GO overlap heavily for
# this dataset and doesn't want the NCBI gene-summary tool voting at all
custom_weights <- list(ncbi_gene_summary = 0, geneset_enrichment = 3)
custom_fusion <- calculate_fusion_score(mm1_packet$fragments, user_weights = custom_weights)
custom_fusion$matrix

# side-by-side comparison -- muting one tool and boosting another shifts
# E_pool/E_evidence without touching the per-fragment magnitude/reliability
# math (.magnitude_score()/.reliability_score() are untouched by tier weight)
data.frame(
    weighting = c('default_tiers', 'custom_user_weights'),
    e_pool = c(default_fusion$e_pool, custom_fusion$e_pool),
    c_dir = c(default_fusion$c_dir, custom_fusion$c_dir),
    e_evidence = c(default_fusion$e_evidence, custom_fusion$e_evidence)
)

# run the same custom_weights through the full synthesis pipeline on this one
# module -- confidence.score and the fusion-string rationale should reflect
# the reweighted E_evidence, and the EVIDENCE CONFIDENCE MATRIX injected into
# the prompt reflects the same custom weights the model was shown
interp_custom_weights <- synthesize_module(
    mm1_packet, desc, backend, user_weights = custom_weights, dataset_context = dataset_ctx
)
interp_custom_weights$confidence$score
interp_custom_weights$confidence$rationale

#---------------------------------------------------------
# export a combined interpretation + evidence HTML report
#---------------------------------------------------------

# pair each module's synthesized narrative with its raw evidence packet and
# compile everything into one browsable HTML file -- open it to eyeball the
# AI narrative against the underlying tool metrics side by side
write_interpretation_report(interps, packets, desc, output_file = 'output/dev_report.html')
```

## Importing pre-existing DEGs

The section below is standalone: it shows how a DEG table you already have
(here generated on the fly with `Seurat::FindMarkers`) is ingested as an
evidence fragment and then fed to the orchestrator alongside the core tools.

```{r deg-generation}
#---------------------------------------------------------
# generate a standard Seurat marker table
#---------------------------------------------------------

# quick contrast on the raw seurat object: Glioblastoma vs the inflammatory
# group, grouped by the diagnosis metadata column. this stands in for any DEG
# table a user already has sitting on disk.
deg_df <- Seurat::FindMarkers(
    seurat_obj,
    ident.1 = 'Glioblastoma',
    ident.2 = 'Inflammatory',
    group.by = 'diagnosis'
)

# inspect the raw Seurat marker frame in the console -- note that features are
# on the rownames and the effect/significance live in avg_log2FC / p_val_adj
head(deg_df)

#---------------------------------------------------------
# ingest the DEG table as an evidence fragment
#---------------------------------------------------------

# no group_col -> a single-contrast cross_condition_delta fragment;
# provenance$source is stamped 'user_supplied' so downstream faithfulness
# checks can tell it apart from a tool-computed fragment
deg_fragment <- import_seurat_markers(
    module_id = 'MM1',
    result = deg_df,
    group_col = NULL
)

# inspect the imported fragment -- compare compact_summary / top_findings here
# against the raw deg_df rows above to confirm the column mapping landed
deg_fragment
deg_fragment$top_findings

#---------------------------------------------------------
# pass the ingested DEGs through the orchestrator
#---------------------------------------------------------

# thin wrapper so the pre-computed DEG table flows through the orchestrator as
# just another tool: it re-tags the imported fragment with whichever module_id
# the orchestrator is currently on, alongside the deterministic tools
deg_import_tool <- function(ctx){
    import_seurat_markers(
        module_id = ctx$module_id,
        result = ctx$params$deg_table,
        group_col = NULL
    )
}

# same tool list as before, now with the DEG import appended
tool_config_deg <- c(tool_config, list(
    list(fn = deg_import_tool, params = list(deg_table = deg_df))
))

packets_deg <- run_orchestrator(ms, tool_config_deg, output_dir = 'output/evidence_packets_deg', modules_use = modules_use)

# inspect one packet -- the ingested DEG fragment should now sit in here next
# to the tool-computed ones
packets_deg[['MM1']]

#---------------------------------------------------------
# synthesize the DEG-augmented packets
#---------------------------------------------------------

# reuse the same desc / backend from the core synthesis block above, now
# running over the packets that carry the imported DEG fragment
interps_deg <- run_synthesis_orchestrator(
    packets_deg, desc, backend, output_dir = 'output/interpretations_deg', dataset_context = dataset_ctx
)

# inspect one interpretation -- check whether the DEG evidence surfaces in the
# synthesized narrative relative to the core-only run
interps_deg[['MM1']]
needs_review(interps_deg[['MM1']])
```

## Pseudo-bulk Analysis Pipeline

The sections below walk `pseudobulk_ModuleSet()` and its two evidence tools
(`differential_module_activity_tool`, `pseudobulk_de_limma_tool`) end to end,
in both usage modes: **standalone** (the pseudo-bulk set runs through
`run_orchestrator()` on its own) and **attached** (the same pseudo-bulk set
rides alongside the real cell-level `ms` from earlier via `with_pseudobulk()`,
so cell- and sample-level tools write into one shared packet per module).
llegir never builds pseudo-bulk itself -- `AggregatePseudobulk()` below is the
same aggregation helper used in `sample_code/T-cell_tumor_network_compare.Rmd`
(sourced from `scripts/pseudobulk_functions.R`; it is not part of the hdWGCNA
package itself, just a project-local aggregation helper), run here on our own
real `seurat_obj`.

```{r pseudobulk-setup}
#---------------------------------------------------------
# aggregate the real seurat_obj counts into a pseudo-bulk SummarizedExperiment
#---------------------------------------------------------

source('scripts/pseudobulk_functions.R')

# one pseudobulk unit per sample (replicate_col == group_col collapses the
# interaction down to sample-level units, same pattern as the sample_code
# T-cell/tumor aggregation); min_cells = 50 keeps every one of our 15 samples,
# the smallest of which carries 106 cells
X <- Seurat::GetAssayData(seurat_obj, layer = 'counts')
meta <- seurat_obj@meta.data
se <- AggregatePseudobulk(X, meta, replicate_col = 'sample', group_col = 'sample', min_cells = 50)

# inspect the raw aggregation before subsetting -- genes x samples counts,
# with per-sample diagnosis/nCells/nUMI riding along in colData
se
SummarizedExperiment::colData(se)[, c('sample', 'diagnosis', 'nCells')]

# differential_module_activity_tool/pseudobulk_de_limma_tool below need a
# clean two-level contrast, so subset down to the same Glioblastoma vs
# Inflammatory comparison used earlier in the DEG-import section, and
# droplevels() so the unused diagnosis categories don't leak into the contrast
se <- se[, SummarizedExperiment::colData(se)$diagnosis %in% c('Glioblastoma', 'Inflammatory')]
SummarizedExperiment::colData(se)$condition <- droplevels(factor(SummarizedExperiment::colData(se)$diagnosis))
table(SummarizedExperiment::colData(se)$condition)

# pull each module's real gene membership + kME straight off the cell-level
# ms -- this is the same module/gene_name/weight shape pseudobulk_ModuleSet()
# expects, just built from the real network instead of a hand-picked subset
pb_genes <- do.call(rbind, lapply(modules_use, function(mod){
    gm <- gene_membership(ms, mod)
    data.frame(module = mod, gene_name = gm$gene_name, weight = gm$kme)
}))
str(pb_genes)
```

```{r pseudobulk-ingest}
#---------------------------------------------------------
# ingest into pseudobulk_ModuleSet() and validate the contract
#---------------------------------------------------------

# pseudobulk_ModuleSet() reads counts/metadata straight off the
# SummarizedExperiment (colData becomes metadata) and re-scores every module
# directly on the real aggregated matrix via decoupleR::run_ulm() -- never by
# averaging the per-cell scores already sitting on ms -- so pb_ms's
# module_scores() are freshly computed pseudo-bulk activity, one row per sample
pb_ms <- pseudobulk_ModuleSet(se, pb_genes, group_col = 'condition', sample_col = 'sample')

# one row per sample, one column per module -- this is the re-scored
# pseudo-bulk activity every downstream tool below reads from
module_scores(pb_ms)

# data_level/aggregated/capabilities read back exactly as the contract expects:
# pseudobulk, already-aggregated, with real counts underneath
pb_ms$data_level
pb_ms$aggregated
capabilities(pb_ms)

# confirms every adapter generic dispatches correctly and the declared
# capabilities/shapes are internally consistent -- throws loudly otherwise
validate_moduleset(pb_ms)
```

```{r pseudobulk-standalone}
#---------------------------------------------------------
# mode 1: standalone pseudo-bulk workflow
#---------------------------------------------------------

# a pseudo-bulk ModuleSet is just another ModuleSet -- hand it straight to
# run_orchestrator() with the two new pseudo-bulk tools; no attachment needed
tool_config_pb <- list(
    list(fn = differential_module_activity_tool, params = list(contrast_col = 'condition')),
    list(fn = pseudobulk_de_limma_tool, params = list(contrast_col = 'condition'))
)

packets_pb_standalone <- run_orchestrator(
    pb_ms, tool_config_pb, output_dir = 'output/evidence_packets_pb_standalone', modules_use = modules_use
)

# MM1-MM3 each get a real module-level differential-activity fragment
# (Glioblastoma vs Inflammatory) plus a real per-gene limma-voom hit --
# compare effect_strength/FDR across the three modules, there is no
# guaranteed winner now that the signal is real rather than simulated
packets_pb_standalone[['MM1']]$fragments[[1]]
packets_pb_standalone[['MM1']]$fragments[[2]]
packets_pb_standalone[['MM2']]$fragments[[1]]
```

```{r pseudobulk-attached}
#---------------------------------------------------------
# mode 2: attach the pseudo-bulk view to the cell-level ms
#---------------------------------------------------------

# with_pseudobulk() stores pb_ms on ms and flips ms's own 'pseudobulk'
# capability to TRUE -- cell-level tools keep reading ms exactly as before,
# and the two pseudo-bulk tools below resolve pb_ms internally via
# pseudobulk_view(), never needing ms and pb_ms passed in separately
ms_with_pb <- with_pseudobulk(ms, pb_ms)
has_capability(ms_with_pb, 'pseudobulk')
identical(pseudobulk_view(ms_with_pb), pb_ms)

# one mixed tool_config: top_genes_tool/cluster_dme_tool read ms_with_pb's own
# cell-level gene_membership/module_scores/metadata; the two pseudo-bulk tools
# transparently pull pb_ms's sample-level view instead -- run_module() never
# needs to know the difference
tool_config_mixed <- list(
    list(fn = top_genes_tool, params = list(n_hubs = 25)),
    list(fn = cluster_dme_tool, params = list(group_by = 'lv2_annot')),
    list(fn = differential_module_activity_tool, params = list(contrast_col = 'condition')),
    list(fn = pseudobulk_de_limma_tool, params = list(contrast_col = 'condition'))
)

packets_mixed <- run_orchestrator(
    ms_with_pb, tool_config_mixed, output_dir = 'output/evidence_packets_pb_mixed', modules_use = modules_use
)

# one packet, four fragments: cell-level top_genes/cluster_dme sit right next
# to the pseudo-bulk differential_module_activity/pseudobulk_de_limma
# fragments -- inspect fragment_id/tool_id across the list to confirm both
# levels landed in the SAME packet
packets_mixed[['MM1']]
vapply(packets_mixed[['MM1']]$fragments, function(f) f$fragment_id, character(1))
```

```{r dataset-context-pb}
#---------------------------------------------------------
# re-run global dataset-level grounding now that a pseudo-bulk view exists
#---------------------------------------------------------

# same dataset_tool_config from the "Global Dataset-Level Grounding" section
# above, now run against ms_with_pb -- dataset_variance_structure_tool can
# resolve pseudobulk_view(ms_with_pb) this time, so both dataset fragments
# land in the context instead of just composition_summary
dataset_ctx_pb <- run_dataset_context(ms_with_pb, dataset_tool_config)
vapply(dataset_ctx_pb$dataset_fragments, function(f) f$tool_id, character(1))
dataset_ctx_pb$dataset_fragments[[2]]$compact_summary
```

```{r pseudobulk-report}
#---------------------------------------------------------
# synthesize the mixed packets and export the combined report
#---------------------------------------------------------

# reuse the same dataset description / backend from the core synthesis block
# above -- now running over packets that carry cell- and sample-level
# evidence side by side, grounded by the fuller dataset_ctx_pb (composition
# AND variance_structure) computed just above
interps_mixed <- run_synthesis_orchestrator(
    packets_mixed, desc, backend, output_dir = 'output/interpretations_pb_mixed', dataset_context = dataset_ctx_pb
)

# inspect one interpretation -- check whether the pseudo-bulk evidence
# (differential_module_activity / pseudobulk_de_limma) surfaces in the
# narrative alongside the cell-level cluster_dme/top_genes evidence
interps_mixed[['MM1']]
needs_review(interps_mixed[['MM1']])

# a single browsable HTML file pairing each module's synthesized narrative
# with its full evidence packet -- open it and confirm the cell-level and
# pseudo-bulk fragments both render in the same module's evidence table
write_interpretation_report(interps_mixed, packets_mixed, desc, output_file = 'output/dev_report_pseudobulk.html')
```
