---
title: 'llegir interactive workflow test'
output: html_document
---

Source/run this line-by-line in an interactive R session to walk the core
package workflow end-to-end on three modules (MM1-MM3). `conda activate
hdWGCNA` first, then step through each chunk and inspect the intermediate
objects in the console as prompted.

```{r setup}
#---------------------------------------------------------
# setup environment and load the dataset
#---------------------------------------------------------

devtools::load_all()

# load the seurat object
data_path <- 'data/CSF_Myeloid_hdWGCNA.rds'
seurat_obj <- readRDS(data_path)

# define the set of modules to use for this test
modules_use <- c('MM1', 'MM2', 'MM3')
```

```{r custom-tool}
#---------------------------------------------------------
# define a custom, script-local evidence tool (Milestone 2 registry demo)
#---------------------------------------------------------

# this is a fully user-defined tool that lives outside the package. it obeys the
# same contract as the package-native tools: it takes a single tool ctx
# (list(ms, module_id, params), built by run_orchestrator) and returns a
# validated evidence_fragment. `ms` and `n_hubs` arrive via ctx$ms /
# ctx$params$n_hubs -- surfaced exactly like hub_genes_tool does it -- so this
# slots straight into tool_config with no orchestrator changes.

# small helper: hit the NCBI E-utils gene endpoints for one symbol and return a
# one-line human summary. esearch resolves the human gene symbol to a UID, then
# esummary pulls the curated description. wrapped by the caller in tryCatch so a
# network failure degrades to the offline mock rather than aborting the run.
.ncbi_summary_one <- function(symbol){

    Sys.sleep(0.4)

    esearch_url <- paste0(
        'https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi',
        '?db=gene&retmode=json&term=', utils::URLencode(symbol),
        '%5Bsym%5D+AND+human%5Borgn%5D'
    )
    uid <- jsonlite::fromJSON(esearch_url)$esearchresult$idlist[1]
    if (is.na(uid) || length(uid) == 0) return(NA_character_)

    esummary_url <- paste0(
        'https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esummary.fcgi',
        '?db=gene&retmode=json&id=', uid
    )
    rec <- jsonlite::fromJSON(esummary_url)$result[[uid]]

    # NCBI's short blurb ('summary') is richer but often empty; fall back to the
    # one-line 'description' so every resolved gene still gets a non-empty string
    summary_text <- rec$summary
    if (is.null(summary_text) || !nzchar(summary_text)) summary_text <- rec$description
    summary_text %||% NA_character_
}

# vectorised lookup with an offline-safe fallback: try the live E-utils call for
# the whole gene vector, and if anything fails (no network, throttling, budget
# lockdown) synthesise a deterministic placeholder so the tool still emits a
# well-formed fragment offline
ncbi_gene_summaries <- function(genes){
    live <- tryCatch(
        vapply(genes, .ncbi_summary_one, character(1)),
        error = function(e) NULL
    )
    if (is.null(live)) {
        return(paste0('[mock summary] ', genes, ': human protein-coding gene (offline placeholder).'))
    }

    # patch only the genes E-utils could not resolve, keep real hits intact
    live[is.na(live)] <- paste0(
        '[mock summary] ', genes[is.na(live)], ': no NCBI record resolved (offline placeholder).'
    )
    unname(live)
}

# the tool proper -- same shape as hub_genes_tool: rank the module's genes by
# kME, keep the top n_hubs, and attach an NCBI one-liner to each
ncbi_gene_summary_tool <- function(ctx){
    n_hubs <- ctx$params$n_hubs %||% 5

    gm <- gene_membership(ctx$ms, ctx$module_id)
    top <- utils::head(gm, n_hubs)
    top$summary <- ncbi_gene_summaries(top$gene_name)

    top_findings <- lapply(seq_len(nrow(top)), function(i){
        list(gene = top$gene_name[i], kme = top$kme[i], summary = top$summary[i])
    })

    compact_summary <- paste0(
        'NCBI summaries for top ', nrow(top), ' hub genes: ',
        paste(top$gene_name, collapse = ', ')
    )

    evidence_fragment(
        fragment_id = 'ncbi_gene_summary',
        tool_id = 'ncbi_gene_summary',
        module_id = ctx$module_id,
        type = 'ranked_genes',
        result = top,
        compact_summary = compact_summary,
        top_findings = top_findings,
        effect_strength = if (nrow(top) > 0) max(top$kme) else 0,
        direction = 'na',
        provenance = make_provenance(
            tool_version = '0.1',
            params = list(n_hubs = n_hubs),
            pkg_versions = pkg_versions(ctx$ms)
        )
    )
}

```

```{r moduleset}

# initialize the ModuleSet adapter
ms <- hdWGCNA_ModuleSet(seurat_obj)
modules(ms)

# run the capabilities check
capabilities(ms)

# smoke-test the script-local custom tool standalone on MM1 before wiring it into
# the orchestrator -- inspect result / top_findings and confirm each gene carries
# a summary string (live NCBI text, or a '[mock summary]' placeholder if offline)
ncbi_demo_frag <- ncbi_gene_summary_tool(list(ms = ms, module_id = 'MM1', params = list(n_hubs = 5)))
ncbi_demo_frag$result
ncbi_demo_frag$top_findings
validate_evidence_fragment(ncbi_demo_frag)

#---------------------------------------------------------
# run core evidence tools over MM1-MM3 only
#---------------------------------------------------------

hallmark_gmt <- c(Hallmark = 'data/h.all.v2026.1.Hs.symbols.gmt')

# define the list of tools to run -- the last entry is the script-local
# ncbi_gene_summary_tool injected right alongside the package-native tools, which
# is all it takes for the registry/orchestrator to treat it as a first-class tool
tool_config <- list(
    list(fn = hub_genes_tool, params = list(n_hubs = 30)),
    list(fn = cluster_dme_tool, params = list(group_by = 'lv2_annot')),
    list(fn = geneset_enrichment_tool, params = list(
        n_hubs = 25,
        db_files = c(GO_BP = 'data/GO_Biological_Process_2026.txt', hallmark_gmt)
    )),
   # list(fn = signature_correlation_tool, params = list(library_files = hallmark_gmt)), # commented this out for now because it takes a long time to run
    list(fn = ncbi_gene_summary_tool, params = list(n_hubs = 5))
)

# run the orchestrator over the list of tools to construct evidence packets per module
packets <- run_orchestrator(ms, tool_config, output_dir = 'output/evidence_packets', modules_use = modules_use)

# inspect one packet's fragments before moving on
packets[['MM1']]
packets[['MM1']]$fragments[[1]]

#---------------------------------------------------------
# run synthesis on the same three modules via GitHub Models
#---------------------------------------------------------

desc <- dataset_description(
    species = 'human',
    tissue = 'cerebrospinal fluid (CSF)',
    cell_compartment = 'myeloid cells',
    assay = 'single-cell RNA-seq (10x)',
    conditions = c(
        'Glioblastoma', 'Brain Metastasis', 'Primary CNS lymphoma',
        'Secondary CNS lymphoma', 'Inflammatory / other neuroinflammatory'
    ),
    notes = 'Modules are CSF-myeloid co-expression programs; interpret in a CNS-myeloid, neuro-oncology / neuroinflammation context.'
)

backend <- cached_backend(
    resolve_backend('github'), provider = 'github', model = .default_models[['github']],
    prompt_template_version = PROMPT_TEMPLATE_VERSION
)

interps <- run_synthesis_orchestrator(packets, desc, backend, output_dir = 'output/interpretations')

# inspect one interpretation
interps[['MM1']]
needs_review(interps[['MM1']])

#---------------------------------------------------------
# export a combined interpretation + evidence HTML report
#---------------------------------------------------------

# pair each module's synthesized narrative with its raw evidence packet and
# compile everything into one browsable HTML file -- open it to eyeball the
# AI narrative against the underlying tool metrics side by side
write_interpretation_report(interps, packets, desc, output_file = 'output/dev_report.html')
```

## Importing pre-existing DEGs

The section below is standalone: it shows how a DEG table you already have
(here generated on the fly with `Seurat::FindMarkers`) is ingested as an
evidence fragment and then fed to the orchestrator alongside the core tools.

```{r deg-generation}
#---------------------------------------------------------
# generate a standard Seurat marker table
#---------------------------------------------------------

# quick contrast on the raw seurat object: Glioblastoma vs the inflammatory
# group, grouped by the diagnosis metadata column. this stands in for any DEG
# table a user already has sitting on disk.
deg_df <- Seurat::FindMarkers(
    seurat_obj,
    ident.1 = 'Glioblastoma',
    ident.2 = 'Inflammatory',
    group.by = 'diagnosis'
)

# inspect the raw Seurat marker frame in the console -- note that features are
# on the rownames and the effect/significance live in avg_log2FC / p_val_adj
head(deg_df)

#---------------------------------------------------------
# ingest the DEG table as an evidence fragment
#---------------------------------------------------------

# no group_col -> a single-contrast cross_condition_delta fragment;
# provenance$source is stamped 'user_supplied' so downstream faithfulness
# checks can tell it apart from a tool-computed fragment
deg_fragment <- import_seurat_markers(
    module_id = 'MM1',
    result = deg_df,
    group_col = NULL
)

# inspect the imported fragment -- compare compact_summary / top_findings here
# against the raw deg_df rows above to confirm the column mapping landed
deg_fragment
deg_fragment$top_findings

#---------------------------------------------------------
# pass the ingested DEGs through the orchestrator
#---------------------------------------------------------

# thin wrapper so the pre-computed DEG table flows through the orchestrator as
# just another tool: it re-tags the imported fragment with whichever module_id
# the orchestrator is currently on, alongside the deterministic tools
deg_import_tool <- function(ctx){
    import_seurat_markers(
        module_id = ctx$module_id,
        result = ctx$params$deg_table,
        group_col = NULL
    )
}

# same tool list as before, now with the DEG import appended
tool_config_deg <- c(tool_config, list(
    list(fn = deg_import_tool, params = list(deg_table = deg_df))
))

packets_deg <- run_orchestrator(ms, tool_config_deg, output_dir = 'output/evidence_packets_deg', modules_use = modules_use)

# inspect one packet -- the ingested DEG fragment should now sit in here next
# to the tool-computed ones
packets_deg[['MM1']]

#---------------------------------------------------------
# synthesize the DEG-augmented packets
#---------------------------------------------------------

# reuse the same desc / backend from the core synthesis block above, now
# running over the packets that carry the imported DEG fragment
interps_deg <- run_synthesis_orchestrator(packets_deg, desc, backend, output_dir = 'output/interpretations_deg')

# inspect one interpretation -- check whether the DEG evidence surfaces in the
# synthesized narrative relative to the core-only run
interps_deg[['MM1']]
needs_review(interps_deg[['MM1']])
```
